---
author: 'Anton Knude Sørensen'
title: 'Text mining, sentiment analysis, and visualization GOT'
date: 'Created on 25 March 2026'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)

library(tidyverse)
library(here)

# For text mining:

library(pdftools)
library(tidytext)
library(textdata) 
library(ggwordcloud)
```

### Get the GOT text:
```{r get-document}
GOT_path <- here("data","got.pdf")
GOT_text <- pdf_text(GOT_path)
```

### Some wrangling:

- Split up pages into separate lines (separated by `\n`) using `stringr::str_split()`
- Unnest into regular columns using `tidyr::unnest()`
- Remove leading/trailing white space with `stringr::str_trim()`

```{r split-lines}
GOT_df <- data.frame(GOT_text) %>% 
  mutate(text_full = str_split(GOT_text, pattern = '\n')) %>% 
  unnest(text_full) %>% 
  mutate(text_full = str_trim(text_full))

# More information: https://cran.r-project.org/web/packages/stringr/vignettes/regular-expressions.html
```

Now each line, on each page, is its own row, with extra starting & trailing spaces removed. 

### Get the tokens (individual words) in tidy format

Use `tidytext::unnest_tokens()` (which pulls from the `tokenizer`) package, to split columns into tokens. We are interested in *words*, so that's the token we'll use:

```{r tokenize}
GOT_tokens <- GOT_df %>% 
  unnest_tokens(word, text_full)

# Each word has its own row!
```

Let's count the words!
```{r count-words}
GOT_wc <- GOT_tokens %>% 
  count(word) %>% 
  arrange(-n)
GOT_wc
```

OK...so we notice that a whole bunch of things show up frequently that we might not be interested in ("a", "the", "and", etc.). These are called *stop words*. Let's remove them. 

### Remove stop words:

See `?stop_words` and `View(stop_words)`to look at documentation for stop words lexicons.

We will *remove* stop words using `tidyr::anti_join()`:
```{r stopwords}
GOT_stop <- GOT_tokens %>% 
  anti_join(stop_words) %>% 
  select(-GOT_text)
```

Now check the counts again: 
```{r count-words2}
GOT_swc <- GOT_stop %>% 
  count(word) %>% 
  arrange(-n)
```

Getting rid of all the numbers (non-text) in `GOT_stop`
```{r skip-numbers}
# This code will filter out numbers by asking:
# If you convert to as.numeric, is it NA (meaning those words)?
# If it IS NA (is.na), then keep it (so all words are kept)
# Anything that is converted to a number is removed

GOT_no_numeric <- GOT_stop %>% 
  filter(is.na(as.numeric(word)))
```

### A word cloud of GOT text words (non-numeric)

See more: https://cran.r-project.org/web/packages/ggwordcloud/vignettes/ggwordcloud.html

```{r wordcloud-prep}
# There are more than 11000 unique words 
length(unique(GOT_no_numeric$word))

# We probably don't want to include them all in a word cloud. Let's filter to only include the top 100 most frequent?
GOT_top100 <- GOT_no_numeric %>% 
  count(word) %>% 
  arrange(-n) %>% 
  head(100)
GOT_top100
```

```{r wordcloud}
GOT_cloud <- ggplot(data = GOT_top100, aes(label = word)) +
  geom_text_wordcloud() +
  theme_minimal()

GOT_cloud
```

That's underwhelming. Let's customize it a bit:
```{r wordcloud-pro}
ggplot(data = GOT_top100, aes(label = word, size = n)) +
  geom_text_wordcloud_area(aes(color = n), shape = "cloud") +
  scale_size_area(max_size = 12) +
  scale_color_gradientn(colors = c("darkgreen","blue","red")) +
  theme_minimal()
```

Cool! And you can facet wrap (for different reports, for example) and update other aesthetics. See more here: https://cran.r-project.org/web/packages/ggwordcloud/vignettes/ggwordcloud.html

### Sentiment analysis

First, check out the ‘sentiments’ lexicon. From Julia Silge and David Robinson (https://www.tidytextmining.com/sentiment.html):

“The three general-purpose lexicons are

  -  AFINN from Finn Årup Nielsen,
  -  bing from Bing Liu and collaborators, and
  -  nrc from Saif Mohammad and Peter Turney

All three of these lexicons are based on unigrams, i.e., single words. These lexicons contain many English words and the words are assigned scores for positive/negative sentiment, and also possibly emotions like joy, anger, sadness, and so forth. The AFINN lexicon assigns words with a score that runs between -5 and 5, with negative scores indicating negative sentiment and positive scores indicating positive sentiment.  The bing lexicon categorizes words in a binary fashion into positive and negative categories. The nrc lexicon categorizes words in a binary fashion (“yes”/“no”) into categories of positive, negative, anger, anticipation, disgust, fear, joy, sadness, surprise, and trust.  All of this information is tabulated in the sentiments dataset, and tidytext provides a function get_sentiments() to get specific sentiment lexicons without the columns that are not used in that lexicon."

Let's explore the sentiment lexicons. "bing" is included, other lexicons ("afinn", "nrc", "loughran") you'll be prompted to download.

"afinn": Words ranked from -5 (very negative) to +5 (very positive)
```{r afinn}
get_sentiments(lexicon = "afinn")
# You should be prompted to install lexicon - choose yes!

```

bing: binary, "positive" or "negative"
```{r bing}
get_sentiments(lexicon = "bing")
# You should be prompted to install lexicon - choose yes!

# Binding words in `GOT_stop` to `bing` lexicon
GOT_bing <- GOT_stop %>% 
  inner_join(get_sentiments("bing"))

#Finding counts by sentiment ranking and plotting them
GOT_bing_hist <- GOT_bing %>% 
  count(sentiment)

# Plot them: 
ggplot(data = GOT_bing_hist, aes(x = sentiment, y = n)) +
  geom_col(aes(fill = sentiment)) +
  theme_bw() +
  labs(title = "Sentiment overview using bing lexicon")


```

nrc:https://saifmohammad.com/WebPages/NRC-Emotion-Lexicon.htm
Includes bins for 8 emotions (anger, anticipation, disgust, fear, joy, sadness, surprise, trust) and positive / negative. 

**Citation for NRC lexicon**: Crowdsourcing a Word-Emotion Association Lexicon, Saif Mohammad and Peter Turney, Computational Intelligence, 29 (3), 436-465, 2013.

Now nrc:
```{r nrc}
get_sentiments(lexicon = "nrc")
# You should be prompted to install lexicon - choose yes!

```

Let's do sentiment analysis on the GOT text data using afinn, and nrc. 


### Sentiment analysis with afinn: 

First, bind words in `GOT_stop` to `afinn` lexicon:
```{r bind-afinn}
GOT_afinn <- GOT_stop %>% 
  inner_join(get_sentiments("afinn"))
GOT_afinn
```

Let's find some counts (by sentiment ranking):
```{r count-afinn}
GOT_afinn_hist <- GOT_afinn %>% 
  count(value)

# Plot them: 
ggplot(data = GOT_afinn_hist, aes(x = value, y = n)) +
  geom_col(aes(fill = value)) +
  theme_bw() +
  labs(title = "Sentiment overview using afinn lexicon")
```

Summarize sentiment for the report: 
```{r summarize-afinn}
GOT_summary <- GOT_afinn %>% 
  summarize(
    mean_score = mean(value),
    median_score = median(value)
  )
GOT_summary 
```

The mean and median indicate *slightly* negative overall sentiments based on the AFINN lexicon. 


### NRC lexicon for sentiment analysis

We can use the NRC lexicon to start "binning" text by the feelings they're typically associated with. As above, we'll use inner_join() to combine the GOT non-stopword text with the nrc lexicon: 

```{r bind-bing}
GOT_nrc <- GOT_stop %>% 
  inner_join(get_sentiments("nrc"))
```

Wait, won't that exclude some of the words in our text? YES! We should check which are excluded using `anti_join()`:

```{r check-exclusions}
GOT_exclude <- GOT_stop %>% 
  anti_join(get_sentiments("nrc"))

# View(GOT_exclude)

# Count to find the most excluded:
GOT_exclude_n <- GOT_exclude %>% 
  count(word, sort = TRUE)

head(GOT_exclude_n)
```

**Lesson: always check which words are EXCLUDED in sentiment analysis using a pre-built lexicon! **

Now find some counts: 
```{r count-bing}
GOT_nrc_n <- GOT_nrc %>% 
  count(sentiment, sort = TRUE)

# And plot them:

ggplot(data = GOT_nrc_n, aes(x = sentiment, y = n)) +
  geom_col(aes(fill = sentiment))+
  theme_bw() +
  labs(title = "Sentiment overview using nrc lexicon")
```

Or count by sentiment *and* word, then facet:
```{r count-nrc}
GOT_nrc_n5 <- GOT_nrc %>% 
  count(word,sentiment, sort = TRUE) %>% 
  group_by(sentiment) %>% 
  top_n(5) %>% 
  ungroup()

GOT_nrc_gg <- ggplot(data = GOT_nrc_n5, aes(x = reorder(word,n), y = n, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~sentiment, ncol = 3, scales = "free") +
  coord_flip() +
  theme_minimal() +
  labs(x = "Word", y = "count")

# Show it
GOT_nrc_gg

# Save it
ggsave(plot = GOT_nrc_gg, 
       here("figures","GOT_nrc_sentiment.png"), 
       height = 8, 
       width = 5)

```
